Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
3 wip

\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{environment}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{language}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\VEC}[1]{#1\k{[]}\xspace}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}

@@

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \null         & \it unit type\\
  & |   & \any          & \it top type\\
  & |   & \environment  & \it environment type \\
  & |   & \Language     & \it language type \\
  & |   & \symbol       & \it symbol type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T & \it function type \\
  & |   & \STRUCT{F_1, ..., F_n} & \it struct type \\
  & |   & \LIST{T}      & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}& \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T & \it arguments \\
  & |   & \VARGS \vspace{5pt} \\
\F & ::= & \ID\, :\,T & \it struct field\vspace{5pt}\\
\V & ::= & ~\VEC\S &\\
  & |   & \NAVEC\S & \it na vector types \vspace{5pt}\\
\S & ::= & \int & \\
  & |   & \chr \\
  & |   & \dbl \\
  & |   & \lgl \\
  & |   & \clx \\
  & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

\subsection{Basic Types}

The workhorse types in R are the vectorized primitives: integers, doubles,
complex numbers, logicals (booleans), characters, and raws (bytes).  These
are vectorized in that they are always considered to be vectors by the R
runtime: for instance, even ``scalar'' numbers are considered to be
unit-length vectors to R.  These primitive vectors are homogeneous, in that
a vector of doubles contains only doubles.

\subsubsection{Possibly-NA Primitives}

scalars and often do high-level
dimensionality checks on their data.  To illustrate, consider the code in
Figure~\ref{fig:scalar-vector-example}.

@@

@@

%
\subsection{Classes}

@@

%
\subsection{Nullable Types}

>>>>

\renewcommand{\null}{\texttt{\textbf{null}}\xspace}
\newcommand{\any}{\texttt{\textbf{any}}\xspace}
\newcommand{\environment}{\texttt{\textbf{environment}}\xspace}
\newcommand{\expression}{\texttt{\textbf{expression}}\xspace}
\newcommand{\Language}{\texttt{\textbf{language}}\xspace} %% Upper case!
\newcommand{\externalptr}{\texttt{\textbf{externalptr}}\xspace}
\renewcommand{\symbol}{\texttt{\textbf{symbol}}\xspace}
\newcommand{\pairlist}{\texttt{\textbf{pairlist}}\xspace}
\newcommand{\weakref}{\texttt{\textbf{weakref}}\xspace}
\renewcommand{\int}{\texttt{\textbf{int}}\xspace}
\newcommand{\chr}{\texttt{\textbf{chr}}\xspace}
\newcommand{\dbl}{\texttt{\textbf{dbl}}\xspace}
\newcommand{\lgl}{\texttt{\textbf{lgl}}\xspace}
\newcommand{\clx}{\texttt{\textbf{clx}}\xspace}
\newcommand{\raw}{\texttt{\textbf{raw}}\xspace}
\newcommand{\NA}{\texttt{{NA}}\xspace}
\newcommand{\VARGS}{\texttt{\textbf{...}}\xspace}
\newcommand{\T}{\ensuremath{T}\xspace}
\renewcommand{\S}{\ensuremath{S}\xspace}
\newcommand{\V}{\ensuremath{V}\xspace}
\newcommand{\A}{\ensuremath{A}\xspace}
\newcommand{\ID}{\ensuremath{ID}\xspace}
\newcommand{\F}{\ensuremath{F}\xspace}
\newcommand{\B}{\ensuremath{B}\xspace}
\newcommand{\FUN}[2]{\ensuremath{\langle #1 \rangle \rightarrow #2}\xspace}
\newcommand{\STRUCT}[1]{\texttt{\textbf{struct}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\LIST}[1]{\texttt{\textbf{list}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\CLASS}[1]{\texttt{\textbf{class}}\ensuremath{\langle #1\rangle}\xspace}
\newcommand{\VEC}[1]{#1\k{[]}\xspace}
\newcommand{\NAVEC}[1]{\k{\string^}\!#1\k{[]}\xspace}

@@

\begin{figure}[!h] \noindent \small  \centering\begin{minipage}{.45\linewidth}
\begin{tabular}{lclr}
\T& ::= & \any          & \it top type\\
  & |   & \null         & \it null type\\
  & |   & \environment  & \it environment type \\
  & |   & \Language     & \it language type \\
  & |   & \symbol       & \it symbol type \\
  & |   & \S            & \it scalar type \\
  & |   & \V            & \it vector type \\
  & |   & \T~\k{|}\T    & \it union type \\
  & |   & \k{?} \T      & \it nullable type \\
  & |   & \FUN{A_1, ..., A_n}\T    & \it function type \\
  & |   & \STRUCT{F_1, ..., F_n}   & \it struct type \\
  & |   & \LIST\T                  & \it list type\\
  & |   & \CLASS{ID_1, ..., ID_n}  & \it class type\vspace{5pt}\\
\end{tabular}\end{minipage}   \hfill
\begin{minipage}{.45\linewidth}\begin{tabular}{lclr}
\A & ::= & \T          & \it arguments \\
   & |   & \VARGS      & \vspace{5pt} \\
\F & ::= & \ID:~~\T & \it struct field\vspace{5pt}\\
\V & ::= & ~\VEC\S     & \\
   & |   & \NAVEC\S    & \it na vector types \vspace{5pt}\\
\S & ::= & \int        & \\
   & |   & \chr \\
   & |   & \dbl \\
   & |   & \lgl \\
   & |   & \clx \\
   & |   & \raw \vspace{5pt}\\
\end{tabular}\end{minipage}\caption{The R type language}\label{types}
\end{figure}

\paragraph{Scalar vs. Vectors} Our type system distinguishes between
vectors of length 1, and vectors of any dimensions. The six base types,
integers, doubles, complex numbers, logicals (booleans), characters, and
raws (bytes), can be either vectors (e.g., \VEC\int) or scalars (e.g.,
\int).  It is worthwhile to note that a vector can happen to be of length 1
and thus a scalar value is a subtype of the corresponding vector type.
Vectors are homogeneous, in that a vector of doubles contains only doubles.

\paragraph{Missing values} In R, a missing value is written
\NA, each of basic types has its specific representation of \NA. Thus there
is an \int missing observation as well as a double missing, and they are
different.  The type system allows to distinguish between vectors that may
contain missing observations (written \NAVEC\int) and those who are
guarantee to be \NA-free (\VEC\int).  A \NA-free vector can be treated as a
vector that may have missing observations.  We do not allow scalar values to
be missings.

\paragraph{Lists} Heterogeneous collections are implemented using lists.
Lists and vectors are closely related. Any vector can be converted to a list
with \code{as.list}, and lists to vectors with \code{unlist} (coercions may
ensue). We support parametrization of the list type, \LIST\T, but for
heterogenous lists this type will default to \any.

\paragraph{Environments}

@@

%
%
\paragraph{Structs}

@@

%
\paragraph{Classes}

@@

%
\paragraph{Nullable Types}
